---
title: "ETL, MDPI & FA"
author: "GregoryH"
date: "`r Sys.Date()`"
output: html_document
---

```{r dependencies, echo = FALSE, include = FALSE}
# Installing Required Packages
# Uncomment the lines below to install packages if they are not already installed.

#install.packages('tidycensus')
#install.packages('waffle')
#install.packages('spdep')

# Load the necessary libraries
library(tidyverse)
library(tidycensus)
library(ggplot2)
library(dplyr)
library(waffle)
library(patchwork)
library(psych)
library(reshape2)
library(colorspace)
library(here)
library(sf)
library(spdep)

# Uncheck to set your Census API key if this is your first time utilizing the API
# census_api_key("KEY HERE")

```

## ETL

This segment specifies the variables that we need to construct the deprivation matrix, the data is then loaded via the get_acs() method.

```{r etl, echo = FALSE, include = FALSE}

# Define ACS variables for each indicator of deprivation
variables_deprivation <- c(
  # Income
  ## Percentage below poverty level
  poverty_total = "B17001_001",
  total_below_poverty = "B17001_002",
  #Total_Pop = "B01003_001",
  
  ## Percentage receiving public assistance 
  recieve_pa_total = "B19057_001",
  total_recieve_pa = "B19057_002",
  
  # Education
  ## Percentage of kids not in school
  kids_Not_In_School = "B14001_010",
  kids_total = "B14001_001",
  
  ## Youth Not Attending School and Not Working
  youth_total = "B14005_001",
  youth_male_xschool_unemployed = "B14005_014",
  youth_male_xschool_xlf = "B14005_015",
  youth_female_xschool_unemployed = "B14005_028",
  youth_female_xschool_xlf = "B14005_029",
  
  ## Percentage of adults 25 above that did not finish schooling
  total_abv25 = "B15003_001",
  abv25_no_school = "B15003_002",
  abv25_nursery = "B15003_003",
  abv25_kindergarten = "B15003_004",
  abv25_g1 = "B15003_005",
  abv25_g2 = "B15003_006",
  abv25_g3 = "B15003_007",
  abv25_g4 = "B15003_008",
  abv25_g5 = "B15003_009",
  abv25_g6 = "B15003_010",
  abv25_g7 = "B15003_011",
  abv25_g8 = "B15003_012",
  abv25_g9 = "B15003_013",
  abv25_g10 = "B15003_014",
  abv25_g11 = "B15003_015",
  abv25_g12 = "B15003_016",
  
  # Health
  ## Disability Status
  dis_total = "B18101_001",
  dis_male_4 = "B18101_004",
  dis_male_5_17 = "B18101_007",
  dis_male_18_34 = "B18101_010",
  dis_male_35_64 = "B18101_013",
  dis_male_65_74 = "B18101_016",
  dis_male_75 = "B18101_019",
  dis_female_4 = "B18101_023",
  dis_female_5_17 = "B18101_026",
  dis_female_18_34 = "B18101_029",
  dis_female_35_64 = "B18101_032",
  dis_female_65_74 = "B18101_035",
  dis_female_75 = "B18101_038",
  
  ## Insurance
  # Placeholder for health insurance variables - to be completed
  # B27001_011,B27001_014,B27001_017,B27001_020,B27001_023,B27001_026
  ins_total = "B18135_001",
  no_ins_dis_18 = "B18135_007",
  no_ins_xdis_18 = "B18135_012",
  no_ins_dis_19_64 = "B18135_018",
  no_ins_xdis_19_64 = "B18135_023",
  no_ins_dis_65 = "B18135_029",
  no_ins_xdis_65 = "B18135_034",
    
  # Housing
  Overcrowded_Housing_own = "B25014_007",
  Overcrowded_Housing_rent = "B25014_013",
  housing_total = "B25002_001",
  cost_burdened_housing_total = "B25070_001",
  cost_burdened_housing_30_35 = "B25070_007",
  cost_burdened_housing_35_40 = "B25070_008",
  cost_burdened_housing_40_50 = "B25070_009",
  cost_burdened_housing_50_above = "B25070_010",
  
  # Living Conditions
  ## Percent without access to <Facility>
  no_plumbing = "B25047_003",
  total_housing_plumbing = "B25047_001",
  no_kitchen_own = "B25053_004",
  no_kitchen_rent = "B25053_007",
  total_housing_kitchen = "B25053_001",
  
  # Digital Divide
  ## Percent without access to <Technology/Infrastructure>
  no_computer = "B28001_011",
  total_households = "B28001_001",
  no_internet = "B28002_013",
  internet_no_subscription = "B28002_012",
  
  # Employment
  ## Percent Unemployed, Percent Not in Labor Force
  unemployed = "B23025_005",
  workforce_total = "B23025_002",
  not_in_labor_force = "B23025_007",
  pop_employment_total = "B23025_001"
  
)

# Pull data for Chicago at the tract level
chicago_data_deprivation <- get_acs(geography = "tract",
                                    variables = variables_deprivation,
                                    state = "IL",
                                    county = "Cook",
                                    output = "wide",
                                    year = 2022,
                                    survey = "acs5")
```

## Data Processing - Building the Deprivation Matrix

The data is then processed to construct the deprivation matrix. In the MDPI literature, Deprivation matrices are often constructed using households or individuals. But since we are constructing a matrix based on census tracts, the percentage of households experiencing deprivation is computed instead.

```{r data_preprocessing, echo = FALSE, include = FALSE}
deprivation_percentages <- chicago_data_deprivation %>%
  mutate(
    # Income
    #Pct_Below_Poverty = (total_below_povertyE / poverty_totalE) * 100, # Blanked out as this should not be in the matrix
    Pct_Receiving_PA = (total_recieve_paE / recieve_pa_totalE) * 100,
    
    # Education
    Pct_Kids_Not_In_School = (kids_Not_In_SchoolE / kids_totalE) * 100,
    Pct_Youth_Not_In_School_Not_Working = ((youth_male_xschool_unemployedE + youth_male_xschool_xlfE + 
                                            youth_female_xschool_unemployedE + youth_female_xschool_xlfE) / youth_totalE) * 100,
    Pct_Adults_No_HighSchool = ((abv25_no_schoolE + abv25_nurseryE + abv25_kindergartenE + abv25_g1E + 
                                 abv25_g2E + abv25_g3E + abv25_g4E + abv25_g5E + abv25_g6E + 
                                 abv25_g7E + abv25_g8E + abv25_g9E + abv25_g10E + abv25_g11E + 
                                 abv25_g12E) / total_abv25E) * 100,
    
    # Health
    Pct_Disabled = ((dis_male_4E + dis_male_5_17E + dis_male_18_34E + dis_male_35_64E + 
                     dis_male_65_74E + dis_male_75E + dis_female_4E + dis_female_5_17E + 
                     dis_female_18_34E + dis_female_35_64E + dis_female_65_74E + 
                     dis_female_75E) / dis_totalE) * 100,
    Pct_No_Insurance = ((no_ins_dis_18E + no_ins_xdis_18E + no_ins_dis_19_64E + 
                         no_ins_xdis_19_64E + no_ins_dis_65E + no_ins_xdis_65E) / ins_totalE) * 100,
    
    # Housing
    Pct_Overcrowded_Housing = ((Overcrowded_Housing_ownE + Overcrowded_Housing_rentE) / housing_totalE) * 100,
    Pct_Cost_Burdened_Housing = ((cost_burdened_housing_30_35E + cost_burdened_housing_35_40E + 
                                  cost_burdened_housing_40_50E + cost_burdened_housing_50_aboveE) / 
                                  cost_burdened_housing_totalE) * 100,
    
    # Living Conditions
    Pct_No_Plumbing = (no_plumbingE / total_housing_plumbingE) * 100,
    Pct_No_Kitchen = ((no_kitchen_ownE + no_kitchen_rentE) / total_housing_kitchenE) * 100,
    
    # Digital Divide
    Pct_No_Computer = (no_computerE / total_householdsE) * 100,
    Pct_No_Internet = (no_internetE / total_householdsE) * 100,
    Pct_Internet_No_Subscription = (internet_no_subscriptionE / total_householdsE) * 100,
    
    # Employment
    Pct_Unemployed = (unemployedE / workforce_totalE) * 100,
    Pct_Not_In_Labor_Force = (not_in_labor_forceE / pop_employment_totalE) * 100
  ) %>%
  select(GEOID, starts_with("Pct_"))  # Keep only GEOID and calculated percentages
```

## Factor Analysis

```{r factor, echo = FALSE, include = FALSE}
deprivation_factors <- deprivation_percentages %>% 
  select(-GEOID) %>%  # FA requires exclusion of GEOID, add back later 
  select_if(is.numeric)

fa_result <- fa(deprivation_factors, nfactors = 5, rotate = "varimax", fm = "minres")

```

### Calculate Poverty rate separately

```{r pov_rate_etl, echo = TRUE, include = FALSE}
variables <- c(Total_Pop = "B01003_001",
               Poverty = "B17001_002")

chicago_tracts_poverty <- get_acs(geography = "tract",
                                  variables = variables,
                                  state = "IL",
                                  county = "Cook",
                                  geometry = TRUE,
                                  output = "wide",
                                  year = 2022,
                                  survey = "acs5")

chicago_tracts_poverty <- chicago_tracts_poverty %>%
  mutate(poverty_rate = (PovertyE / Total_PopE) * 100)
```

Append factor scores to each GEOID and merge with poverty rate

```{r factor_scpres}
# Calculate factor scores for each unit of analysis
factor_scores <- factor.scores(deprivation_percentages[-1], fa_result)$scores  # Excluding 'GEOID' for calculation

# Convert into df
factor_scores_df <- as.data.frame(factor_scores)

# Add 'GEOID' back
factor_scores_df$GEOID <- deprivation_percentages$GEOID

# Renaming the MR factors to align with the op-ed (better readability)
factor_names <- c(
  "Education and Health", 
  "Employment and Shelter", 
  "Labor and Housing Cost", 
  "The Digital Divide", 
  "School Attendance"
)

names(factor_scores_df) <- c(factor_names, "GEOID")

merged_data <- left_join(chicago_tracts_poverty, factor_scores_df, by = "GEOID")
```

## Neighbor averages for more spatial context

```{r compute_neighbor_averages}
# Convert GEOID to character
merged_data$GEOID <- as.character(merged_data$GEOID)
neighbor_list <- st_touches(merged_data, sparse = FALSE)

# Each row in neighbor_list corresponds to a GEOID and contains TRUE/FALSE
# for each other GEOID, indicate whether it is a neighbor
neighbor_indices <- lapply(1:nrow(merged_data), function(i) {
  which(neighbor_list[i, ])
})

# Calculate neighboring averages
factor_names <- c(
  "poverty_rate",
  "Education and Health", 
  "Employment and Shelter", 
  "Labor and Housing Cost", 
  "The Digital Divide", 
  "School Attendance"
)

for (factor_name in factor_names) {
  # Extract the values for the current factor
  factor_values <- merged_data[[factor_name]]

  # Pre-allocate a vector to store the neighbor averages
  neighbor_averages <- vector("numeric", length = nrow(merged_data))
  
  for (i in seq_len(nrow(merged_data))) {
    neighbors_i <- neighbor_indices[[i]]
    
    if (is.null(neighbors_i) || length(neighbors_i) == 0) {
      neighbor_averages[i] <- NA  # Assign NA if there are no neighbors
    } else {
      neighbor_scores <- factor_values[neighbors_i]
      
      if (all(is.na(neighbor_scores))) {
        neighbor_averages[i] <- NA  # Assign NA if all neighbor scores are NA
      } else {
        neighbor_averages[i] <- mean(neighbor_scores, na.rm = TRUE)
      }
    }
  }
  
  # Add new column for neighboring averages
  new_col_name <- paste0(factor_name, "_neighbor_average")
  merged_data[[new_col_name]] <- neighbor_averages
}
```

## Add names of community areas for more spatial context

```{r community_names}
community_data <- read.csv("ct_chicago.csv", stringsAsFactors = FALSE)
merged_data <- merge(merged_data, community_data, by.x = "GEOID", by.y = "ct_chicago", all.x = TRUE)
merged_data <- merged_data %>%
  mutate(community_name = if_else(is.na(community_name), 'Cook County - Outside Chicago', community_name))
```


```{r save_data}
data_dir <- here("interactive_mdpi")
saveRDS(merged_data, file.path(data_dir, "factor_scores_geoids.rds"))
```
