---
title: "Exploring Multidimensional Poverty in the Chicago Area"
---

# Multidimensional Poverty in Cook County

Within the vibrant urban tapestry of Chicago, characterized by its architectural prowess and economic vitality, lies a less visible but critically important dimension of urban life: multidimensional poverty. This phenomenon extends beyond the traditional income-based assessments, to encompass a range of deprivations that affect households.

This commentary draws upon the latest empirical data from the ACS 5-Year Estimates for 2022, employing a methodical approach ([more details here](multidimensional_poverty.qmd)) to identify and quantify 15 distinct indicators of deprivation. These indicators span critical domains such as health, education, employment, housing, and digital accessibility, thereby offering a comprehensive view of the multifaceted nature of poverty.

The rationale for adopting a multidimensional perspective on poverty is grounded in contemporary scholarly discourse, which advocates for a more nuanced understanding of poverty that reflects the complex realities of urban deprivation.

**Consider this - What does poverty really mean in an urban context?**

It's not just about empty pockets; it's also about being deprived in many other areas of life, such as quality education, stable jobs, safe homes, and equitable access in a digitalized world.

## The Chicago Deprivation Canvas

```{r dependencies, echo = FALSE, include = FALSE}
# Installing Required Packages
# packages <- c("tidycensus", "waffle", "ggplot2", "dplyr", "patchwork", "psych", "reshape2", "colorspace")
# sapply(packages, function(p) if(!require(p, character.only = TRUE)) install.packages(p))

# Load the necessary libraries
library(tidyverse)
library(tidycensus)
library(ggplot2)
library(dplyr)
library(waffle)
library(patchwork)
library(psych)
library(reshape2)
library(colorspace)

```

```{r viz1_etl, echo = FALSE, include = FALSE}
# Define ACS variables for each indicator of deprivation
variables_deprivation <- c(
  # Income
  ## Percentage below poverty level
  poverty_total = "B17001_001",
  total_below_poverty = "B17001_002",
  
  ## Percentage receiving public assistance 
  recieve_pa_total = "B19057_001",
  total_recieve_pa = "B19057_002",
  
  # Education
  ## Percentage of kids not in school
  kids_Not_In_School = "B14001_010",
  kids_total = "B14001_001",
  
  ## Youth Not Attending School and Not Working
  youth_total = "B14005_001",
  youth_male_xschool_unemployed = "B14005_014",
  youth_male_xschool_xlf = "B14005_015",
  youth_female_xschool_unemployed = "B14005_028",
  youth_female_xschool_xlf = "B14005_029",
  
  ## Percentage of adults 25 above that did not finish schooling
  total_abv25 = "B15003_001",
  abv25_no_school = "B15003_002",
  abv25_nursery = "B15003_003",
  abv25_kindergarten = "B15003_004",
  abv25_g1 = "B15003_005",
  abv25_g2 = "B15003_006",
  abv25_g3 = "B15003_007",
  abv25_g4 = "B15003_008",
  abv25_g5 = "B15003_009",
  abv25_g6 = "B15003_010",
  abv25_g7 = "B15003_011",
  abv25_g8 = "B15003_012",
  abv25_g9 = "B15003_013",
  abv25_g10 = "B15003_014",
  abv25_g11 = "B15003_015",
  abv25_g12 = "B15003_016",
  
  # Health
  ## Disability Status
  dis_total = "B18101_001",
  dis_male_4 = "B18101_004",
  dis_male_5_17 = "B18101_007",
  dis_male_18_34 = "B18101_010",
  dis_male_35_64 = "B18101_013",
  dis_male_65_74 = "B18101_016",
  dis_male_75 = "B18101_019",
  dis_female_4 = "B18101_023",
  dis_female_5_17 = "B18101_026",
  dis_female_18_34 = "B18101_029",
  dis_female_35_64 = "B18101_032",
  dis_female_65_74 = "B18101_035",
  dis_female_75 = "B18101_038",
  
  ## Insurance
  # Placeholder for health insurance variables - to be completed
  # B27001_011,B27001_014,B27001_017,B27001_020,B27001_023,B27001_026
  ins_total = "B18135_001",
  no_ins_dis_18 = "B18135_007",
  no_ins_xdis_18 = "B18135_012",
  no_ins_dis_19_64 = "B18135_018",
  no_ins_xdis_19_64 = "B18135_023",
  no_ins_dis_65 = "B18135_029",
  no_ins_xdis_65 = "B18135_034",
    
  # Housing
  Overcrowded_Housing_own = "B25014_007",
  Overcrowded_Housing_rent = "B25014_013",
  housing_total = "B25002_001",
  cost_burdened_housing_total = "B25070_001",
  cost_burdened_housing_30_35 = "B25070_007",
  cost_burdened_housing_35_40 = "B25070_008",
  cost_burdened_housing_40_50 = "B25070_009",
  cost_burdened_housing_50_above = "B25070_010",
  
  # Living Conditions
  ## Percent without access to <Facility>
  no_plumbing = "B25047_003",
  total_housing_plumbing = "B25047_001",
  no_kitchen_own = "B25053_004",
  no_kitchen_rent = "B25053_007",
  total_housing_kitchen = "B25053_001",
  
  # Digital Divide
  ## Percent without access to <Technology/Infrastructure>
  no_computer = "B28001_011",
  total_households = "B28001_001",
  no_internet = "B28002_013",
  internet_no_subscription = "B28002_012",
  
  # Employment
  ## Percent Unemployed, Percent Not in Labor Force
  unemployed = "B23025_005",
  workforce_total = "B23025_002",
  not_in_labor_force = "B23025_007",
  pop_employment_total = "B23025_001"
  
)

# Pull data for Chicago at the tract level
chicago_data_deprivation <- get_acs(geography = "tract",
                                    variables = variables_deprivation,
                                    state = "IL",
                                    county = "Cook",
                                    output = "wide",
                                    year = 2022,
                                    survey = "acs5")
```

```{r viz1_processing, echo = FALSE, include = FALSE}

deprivation_percentages <- chicago_data_deprivation %>%
  mutate(
    # Income
    #Pct_Below_Poverty = (total_below_povertyE / poverty_totalE) * 100,
    Pct_Receiving_PA = (total_recieve_paE / recieve_pa_totalE) * 100,
    
    # Education
    Pct_Kids_Not_In_School = (kids_Not_In_SchoolE / kids_totalE) * 100,
    Pct_Youth_Not_In_School_Not_Working = ((youth_male_xschool_unemployedE + youth_male_xschool_xlfE + 
                                            youth_female_xschool_unemployedE + youth_female_xschool_xlfE) / youth_totalE) * 100,
    Pct_Adults_No_HighSchool = ((abv25_no_schoolE + abv25_nurseryE + abv25_kindergartenE + abv25_g1E + 
                                 abv25_g2E + abv25_g3E + abv25_g4E + abv25_g5E + abv25_g6E + 
                                 abv25_g7E + abv25_g8E + abv25_g9E + abv25_g10E + abv25_g11E + 
                                 abv25_g12E) / total_abv25E) * 100,
    
    # Health
    Pct_Disabled = ((dis_male_4E + dis_male_5_17E + dis_male_18_34E + dis_male_35_64E + 
                     dis_male_65_74E + dis_male_75E + dis_female_4E + dis_female_5_17E + 
                     dis_female_18_34E + dis_female_35_64E + dis_female_65_74E + 
                     dis_female_75E) / dis_totalE) * 100,
    Pct_No_Insurance = ((no_ins_dis_18E + no_ins_xdis_18E + no_ins_dis_19_64E + 
                         no_ins_xdis_19_64E + no_ins_dis_65E + no_ins_xdis_65E) / ins_totalE) * 100,
    
    # Housing
    Pct_Overcrowded_Housing = ((Overcrowded_Housing_ownE + Overcrowded_Housing_rentE) / housing_totalE) * 100,
    Pct_Cost_Burdened_Housing = ((cost_burdened_housing_30_35E + cost_burdened_housing_35_40E + 
                                  cost_burdened_housing_40_50E + cost_burdened_housing_50_aboveE) / 
                                  cost_burdened_housing_totalE) * 100,
    
    # Living Conditions
    Pct_No_Plumbing = (no_plumbingE / total_housing_plumbingE) * 100,
    Pct_No_Kitchen = ((no_kitchen_ownE + no_kitchen_rentE) / total_housing_kitchenE) * 100,
    
    # Digital Divide
    Pct_No_Computer = (no_computerE / total_householdsE) * 100,
    Pct_No_Internet = (no_internetE / total_householdsE) * 100,
    Pct_Internet_No_Subscription = (internet_no_subscriptionE / total_householdsE) * 100,
    
    # Employment
    Pct_Unemployed = (unemployedE / workforce_totalE) * 100,
    Pct_Not_In_Labor_Force = (not_in_labor_forceE / pop_employment_totalE) * 100
  ) %>%
  select(GEOID, starts_with("Pct_"))  # Keep only GEOID and calculated percentages
```

```{r factor, echo = FALSE, include = FALSE}
# Exclude non-numeric variables if any
deprivation_factors <- deprivation_percentages %>% 
  select(-GEOID) %>%  # Exclude GEOID or any other non-numeric variables
  select_if(is.numeric)  # Select only numeric columns

fa_result <- fa(deprivation_factors, nfactors = 5, rotate = "varimax", fm = "minres")

```

**Figure 1: Multidimensional Poverty Indicators in Chicago: A Factor Analysis**

```{r viz1, echo=FALSE, include=TRUE, fig.width=14, fig.height=8}
#supress warnings
options(warn = -1)
options(message = -1)
options(tigris_use_cache = TRUE)

# Extract loadings matrix
loadings_matrix <- fa_result$loadings

# Assuming the last 3 rows are SS loadings, prop. var., and cum. var., remove them
loadings_matrix_clean <- loadings_matrix[1:(nrow(loadings_matrix)), ]

# Convert clean loadings matrix to a dataframe
loadings_df <- as.data.frame(loadings_matrix_clean)

# Add row names as a new column for variable names
loadings_df$Variable <- rownames(loadings_df)

# Define the order
Ord <- rev(c("Pct_Adults_No_HighSchool",
         "Pct_No_Insurance",
         "Pct_Unemployed",
         "Pct_No_Plumbing",
         "Pct_Receiving_PA",
         "Pct_Not_In_Labor_Force",
         "Pct_Disabled",
         "Pct_No_Computer",
         "Pct_No_Internet",
         "Pct_Cost_Burdened_Housing",
         "Pct_Kids_Not_In_School",
         "Pct_Internet_No_Subscription",
         "Pct_Overcrowded_Housing",
         "Pct_Youth_Not_In_School_Not_Working",
         "Pct_No_Kitchen"
         ))

# Set the factor levels based on the order
loadings_df$Variable <- factor(loadings_df$Variable, levels = Ord)

# Rename factor columns
colnames(loadings_df)[1:ncol(loadings_df)-1] <- c("Educational and Health",
                                                  "Employment & Shelter",
                                                  "Labor & Housing Cost",
                                                  "Digital Divide",
                                                  "School Attendance")

# Melt dataframe
loadings_long <- melt(loadings_df, id.vars = "Variable", variable.name = "Factor", value.name = "Loading")

# Rename variables for better readability
loadings_long$Variable <- factor(loadings_long$Variable, levels = Ord,
                                 labels = rev(c('Adults 25 above that did not finish schooling',
                                            'Adults without health insurance',
                                            'Unemployed',
                                            'No Plumbing',
                                            'Receiving public assistance',
                                            'Not in Labor Force',
                                            'Disabled',
                                            'No Computer',
                                            'No Internet Access',
                                            'House-Cost burdened',
                                            'Kids not enrolled in school',
                                            'No internet subscription',
                                            'House Overcrowded',
                                            'Youth neither attending school nor work',
                                            'No kitchen facility')))

ggplot(loadings_long, aes(Variable, abs(Loading), fill=Loading)) + 
  facet_wrap(~ Factor, nrow=1) + 
  geom_bar(stat="identity", color="black", size=0.3) +  # change outline to black, increase size
  coord_flip() + 
  scale_fill_continuous_diverging(palette = "Cyan-Mage", l1 = 30, l2 = 100, p1 = .9, p2 = 1.2) +
  labs(x= "Dimension of Deprivation", y = "Loading Strength", fill = "Loading Strength") +
  theme_bw(base_size=10) +
  theme(
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey93", colour = "black", size = 0.5),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 12),
    axis.text.y.left = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    axis.text.x.bottom = element_text(size = 11),
    legend.text = element_text(size = 12),  # Adjust legend text
    legend.title = element_text(size = 14),  # Adjust legend title
    legend.background = element_rect(fill = "grey93", colour = "black", size = 0.5),
    panel.spacing = unit(0.33, "lines"),  # spacing between facets
    plot.margin = margin(12, 12, 12, 12)  # reminder: plot margins (top, right, bottom, left)
  )
```

Figure 1 presents a factor analysis of various deprivation indicators within Chicago, offering a statistical representation of the underlying structures of multidimensional poverty. Factor analysis, a complex statistical method, is employed here to identify latent variables that influence the observed data. These latent variables (or factors), represent broad domains of deprivation that are inferred from the correlations among the individual indicators. By analyzing these relationships, the figure visually conveys which aspects of urban life are most strongly associated with the different facets of poverty.

Each bar on this graph tells a story: a slice of life defined by absence — be it in health insurance, a roof over one's head, or a computer in one's home. These bars stretch out to the extent of their 'factor loading strength,' a numerical representation of how strongly each indicator resonates with a factor (facet of poverty). Longer bars signifies a stronger association with a particular facet.

We've dissected the graph into thematic territories, each named for a facet of deprivation like "Educational and Health," or "Digital Divide." These are not mere labels but analytical realms — Multiple Regression scores that cluster the indicators into meaningful assemblies. Here, the challenges of poverty don't stand in isolation; they cluster and often overlap, offering insight into which factors bundle together in the urban landscape of Chicago.
